---
title: "F6_Set12 EdgeR analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Last modified Apr 27, 2020. Author JWS
This analysis is in part to quickly visualize and assess the dataset. 

### Load Libraries
```{r message=FALSE, warning=FALSE}

library(dplyr) 
library(magrittr)

library(reshape2)
library(ggplot2)

library(knitr)

```


### Upload Data and meta data (library information and geneID/geneName)
```{r}

# Ensenble gene ID + FANTOM CAT ID is used
data = read.table("local data/F6_set12_count.tsv", header=TRUE, row.names=1)

# File matching CAGE library ID with human readable sample names
meta = read.csv("meta/F6_set12_meta.csv")
#meta = dplyr::arrange(meta, perturb_id)

# Match ID and replace with human readable sample names
# order of match function switched from the original script
colnames(data)[ match(meta$library_name, colnames(data)) ] = as.character(meta$perturb_id)

# Upload file to match geneID to geneName
annotations = read.csv("meta/GeneID_to_GeneName.csv", header=TRUE)


```

### Run EdgeR
```{r message=FALSE, warning=FALSE}

library(edgeR)
library(pheatmap)

# Create a GROUP based on header names by removing .1, .2, .3
s = strsplit(names(data), split="\\.")
group = sapply(1:length(s), function(X)  s[[X]][1])

# Standard EdgeR commands for normalization and filtering
my_data = DGEList(counts=data, group=group)
keep = rowSums(cpm(my_data)) >= 3 # can be changed 

my_data = my_data[keep, ]
my_data = calcNormFactors(my_data)

# Plot MDS for QC
plotMDS(my_data)

# Plot correlation matrix
my_data.cor = cor(cpm(my_data), method="pearson")
pheatmap(my_data.cor)

# Boxplot
boxplot(log2(cpm(my_data)), las=2, main="Normalized CPM")

# Create design matrix
design = model.matrix(~0+group)
rownames(design) = colnames(my_data)
colnames(design) = unique(group)

design # based on replicates

# Fit GLM
y <- estimateDisp(my_data, design)
fit <- glmFit(y, design)

# Make comparisons
con <- makeContrasts(Day08 = iN_08d - iN_00d, Day03 = iN_03d - iN_00d, levels=design)

lrt <- glmLRT(fit, contrast = con)

# Show top DEG and summarize
#topTags(lrt)
summary(decideTests(lrt))


```

### Retrieve significantly differentially expressed genes
```{r}

# Rename
de.table = lrt$table

# Perform mulitple testing and add column
FDR <- p.adjust(de.table$PValue, method="BH") 
de.table$FDR = FDR
de.table$geneID = rownames(de.table)

# Match geneID and merge in geneName
de.table = merge(de.table, annotations, by = "geneID")

# Filter rows (genes) based on a defined criteria using dplyr
hm.table = de.table %>% dplyr::filter(., FDR < 0.01, (logFC.Day08 > 1 | logFC.Day08 < (-1))) %>% 
                        dplyr::mutate(., logFC.Day00 = 0) %>% 
                        dplyr::select(geneID, geneName, logFC.Day00, logFC.Day03, logFC.Day08, FDR)

dim(hm.table)

```

### Search for gene of interest and Visualize
```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=4}

# Optional to use unfiltered data
#de.table2 = de.table %>% dplyr::mutate(., logFC.Day00 = 0) %>%  dplyr::select(geneID, geneName, logFC.Day00, logFC.Day03, logFC.Day08, FDR)

# goi = gene of interest
goi = c("MAP2", "ASCL1", "NEUROG2", "PAX6", "SREBF2", "SOX11", "SOX2", "HES2") # Add gene symbols here

goi.table = dplyr::filter(hm.table, geneName %in% goi)

melt.table = dplyr::filter(hm.table, geneName %in% goi) %>% dplyr::select(-FDR) %>% melt()

ggplot(melt.table, aes(x=variable, y=as.numeric(value), color=factor(geneName))) +
                        geom_line(aes(group=factor(geneName)), size=1.2, alpha=0.5) + 
                        geom_point(aes(group=factor(geneName))) + 
                        ylab("Log2-ratio") + xlab ("Days") + theme_light(base_size = 20) +     scale_color_brewer(palette="Dark2", name="") + 
  scale_x_discrete( labels=c("00", "03", "08"))

```


### Perform Gene Ontology anlaysis
```{r message=FALSE, warning=FALSE, cache=TRUE}

library(topGO)

# Define background for GO (also try with hm.table$Symbol) 
universe = de.table$geneID

# Identify sigGenes for day 3
sigGenes.03 = de.table %>% dplyr::filter((logFC.Day03 > 1 | logFC.Day03 < (-1)), 
                                         FDR < 0.05) %>% 
                                      dplyr::select(geneID) %>% .[,1]

# Identify sigGenes for day 8
sigGenes.08 = de.table %>% dplyr::filter((logFC.Day08 > 1 | logFC.Day08 < (-1)), 
                                         FDR < 0.05) %>% 
                                      dplyr::select(geneID) %>% .[,1]

# Convert format to factor
gene_universe.03 = factor(as.numeric(universe %in% sigGenes.03))   
gene_universe.08 = factor(as.numeric(universe %in% sigGenes.08))   
names(gene_universe.03) = universe
names(gene_universe.08) = universe

# Run Top GO for day 3
go_data <- new("topGOdata",
               ontology = "BP",  
               allGenes = gene_universe.03,   
               nodeSize = 5,  
               annot = annFUN.org,  
               mapping = "org.Hs.eg.db", 
               ID = "ensembl")  

# Statistics 
go_test <- runTest(go_data, algorithm = "classic", statistic = "fisher")

# Extract GO terms and create table
go_table <- GenTable(go_data, weightFisher = go_test,
                     orderBy = "weightFisher", ranksOf = "weightFisher",
                     topNodes = sum(score(go_test) < .01), numChar = 50 )


# Run Top GO for day 8
go_data2 <- new("topGOdata",
               ontology = "BP",  
               allGenes = gene_universe.08,   
               nodeSize = 5,  
               annot = annFUN.org,  
               mapping = "org.Hs.eg.db", 
               ID = "ensembl")  

# Statistics 
go_test2 <- runTest(go_data2, algorithm = "classic", statistic = "fisher")

# Extract GO terms and create table
go_table2 <- GenTable(go_data2, weightFisher = go_test2,
                     orderBy = "weightFisher", ranksOf = "weightFisher",
                     topNodes = sum(score(go_test2) < .01), numChar = 50 )


# Merge Day 3 and Day 8 GO tables
go_table_all = merge(go_table, go_table2, by = "GO.ID")

# Change "< 1e-30" character string into a numeric value
go_table_all$weightFisher.x = as.numeric(gsub("< 1e-30", 1e-30, go_table_all$weightFisher.x))
go_table_all$weightFisher.y = as.numeric(gsub("< 1e-30", 1e-30, go_table_all$weightFisher.y))

```


### Plot and visualize Significant GO terms
```{r message=FALSE, warning=FALSE, figure.width=30, fig.height = 7}

# Convert p-values to -log10. 
# .x = day 3
# .y = day 8
go_table_all$weightFisher.x = go_table_all$weightFisher.x %>% as.numeric %>% -log10(.)
go_table_all$weightFisher.y = go_table_all$weightFisher.y %>% as.numeric %>% -log10(.)

# Sort the table based on the GO ID significance on day 8
go_table_all$GO.ID = factor(go_table_all$GO.ID, 
                      levels = dplyr::arrange(go_table_all, desc(weightFisher.y))$GO.ID )

# Keep only the relevant columns and make GO term unique
m = dplyr::select(go_table_all, GO.ID, Term.x, weightFisher.x, weightFisher.y)
m$Term.x = make.unique(m$Term.x)

# Rearrange the table for GGplot
mx = melt(m)

# Rename columns
mx$variable = gsub("weightFisher.x", "Day03", mx$variable)
mx$variable = gsub("weightFisher.y", "Day08", mx$variable)

# Reorder the terms based on P-value
mx$Term.x = factor(mx$Term.x, levels = as.factor(m$Term.x[order(m$weightFisher.y, decreasing = TRUE)]) )

# Extract terms that contain "Neuro"
keep = grep("neuro", mx$Term.x, ignore.case = TRUE)
mx.neuro = mx[keep,]

# Plot GO terms
ggplot(mx.neuro, aes(x=Term.x, y=value, fill = variable)) + geom_bar(stat = "identity", position = "dodge") +
          theme_light(base_size = 13) +  
          theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
          xlab("") + ylab("-log10(p-value)") 


```

